\pkg{sdmTMB}: An \proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields

### 2023-10-19 doc/paper-jss/sdmTMB-paper.Rnw
Edits + fct + code

Data are often collected in space or in space repeatedly over time. While such data are a rich source of information across many fields \citep[e.g.,][]{legendre1989a, rossi1992, tilman1997spatial},
% TODO: other non-ecology refs
they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart due to measured and unmeasured variables \citep{cressie1993, diggle2007, cressie2011}.
While measured variables can be accounted for with predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables can cause residual spatial correlation.
Accounting for this residual correlation is important because doing so allows for valid statistical inference \citep{legendre1989a, dormann2007}, can improve predictions \citep[e.g.,][]{shelton2014}, and can be of interest itself \citep[e.g.,][]{thorson2019d, barnett2021}.

Geostatistical generalized linear mixed effects models (GLMMs) with spatially correlated random effects constitute an appropriate class of models for such data \citep{rue2005gmrf, diggle2007, cressie2011}.
Similarly to how random intercepts can account for correlation among groups, spatial or spatiotemporal random effects can account for unmeasured variables that cause observations to be correlated in space or both space and time.
A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed to be drawn from a multivariate normal distribution, constrained by specific covariance functions such as the exponential or Matérn \citep{cressie1993, diggle2007, chiles1999}.
% TODO: decide on using GRF throughout

@@

A basic GLMM with spatial Gaussian random fields can be written as
\[
\begin{aligned}
\mathbb{E}[y_{\bm{s}}] &= \mu_{\bm{s}},\\
\mu_{\bm{s}} &=
g^{-1} \left( \eta_{\bm{s}} \right),\\
\eta_{\bm{s}} &= \bm{X}_{\bm{s}} \bm{\beta} + \omega_{\bm{s}},\\
\end{aligned}
\]
where the expected value $\mathbb{E}[.]$ of an observation $y$ at coordinates in space $\bm{s}$ is equal to mean $\mu_s$.
That mean is the result of an inverse link function $g^{-1}$ applied to a linear predictor $\eta$.
In this case, that linear predictor is the combination of a model matrix $\bm{X}_{\bm{s}}$ multiplied by a vector of coefficients $\bm{\beta}$ and a value from a spatial random field $\omega_{\bm{s}}$.
This spatial random field represents the effect of latent spatial variables that are not otherwise accounted for in the model.
Alternatively, $\omega_{\bm{s}}$ can be thought of representing spatially correlated noise.
More simply, the vector of $\omega_{\bm{s}}$ ($\bm{\omega}$) represents a ``wiggly'' surface with an expected value of zero that is added to the linear predictor in link space (e.g., Figure~\ref{fig:matern-range}).
The vector $\bm{\omega}$ is assumed drawn from a multivariate normal distribution with a covariance matrix $\boldsymbol{\Sigma}_\omega$ constrained by some function that defines the rate at which spatial covariance decays with distance,
\[
\boldsymbol{\omega} \sim \operatorname{MVNormal}
  \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\omega \right).
\]

@@

Various covariance functions are possible, but a popular and flexible one is the Mat\'ern \citep{whittle1954, matern1986} (Figure~\ref{fig:matern-range}).
If we define $\lVert h \rVert$ as the Euclidean distance between spatial points $\bm{s}_j$ and $\bm{s}_k$, we can represent the Mat\'ern covariance $\Phi$ as
\[
\Phi\left( \bm{s}_j, \bm{s}_k \right) =
  \frac{\sigma^2}{2^{\nu - 1}\Gamma(\nu)}
  \left(\kappa \lVert h \rVert \right)^\nu K_\nu
  \left( \kappa \lVert h \rVert \right).
\]
The parameter $\sigma^2$ is the marginal variance (magnitude of the random field ``wiggliness''), $\Gamma$ represents the Gamma function, $K_\nu$ represents the modified Bessel function of the second kind, and $\kappa$ represents the spatial decorrelation rate.
The parameter $\nu$ controls the smoothness of the covariance function.
In practice, $\nu$ is challenging to estimate and herein is fixed at $\nu = 1$ \citep{lindgren2011}.
A more interpretable parameter than the spatial decorrelation rate $\kappa$ is the spatial range---a distance at which two points are effectively independent.
A common definition is $\sqrt{8 \nu} / \kappa$ (so if $\nu = 1$, $\sqrt{8} / \kappa$), which has the empirically derived property that correlation decays to $\approx 0.13$ \citep{lindgren2011} (Figure~\ref{fig:matern-range}).

@@

In practice, working directly with the dense covariance matrix $\bm{\Sigma}$ is computationally expensive
and methods for working directly with its inverse, the precision matrix $\bm{Q}$, are more efficient ($\bm{Q} = \bm{\Sigma}^{-1}$) \citep{simpson2012}.
One such approach is the SPDE approach.
A full description of the SPDE approach is beyond the scope of this paper.
Instead, we refer to the following literature:
\citet{lindgren2011} develops the approach.
\citet{lindgren2015} and \cite{bakka2018review} summarize the SPDE approach for spatial modelling in the context of \INLA.
The second chapter of \citet{krainski2018} provides an overview of the SPDE approach to spatial modelling with a focus on linking theory to code.
\citet{miller2019} contains an introduction to the SPDE and illustrates its equivalence to penalized smoothing approaches.
\citet{lindgren2022ten} provides a recent review of the approach and its applications over the last decade.

For a user of \sdmTMB, the following are the important elements to understand.
First, the SPDE approach links Gaussian random fields (GRFs) that have a Mat\'ern covariance function to Gaussian \emph{Markov} random fields (GMRFs) in such a way that a GMRF can be a good approximation to a GRF \citep{lindgren2011}.
This means that GRF models can be computationally fit as GMRFs.
By working with GMRFs, one can take advantage of theory developed to estimate their sparse precision matrix efficiently \citep{rue2005gmrf, lindgren2011} and avoid the inversion of large dense matrices.
Second, the SPDE involves piece-wise linear basis functions that are defined by a triangulation over the spatial area of interest \citep{lindgren2011}.
Commonly, this is referred to as a ``mesh''.
The properties of this mesh (e.g., its resolution and how far it extends beyond the data) affect the accuracy and computational efficiency of the SPDE approach \citep{lindgren2011}.
Third, working with the SPDE approach and the precision matrix of a GMRF ($\bm{Q}$) introduces an additional parameter $\tau$, which scales the precision matrix, and involves three sparse matrices associated with the mesh ($\bm{C}$, $\bm{G}_1$, and $\bm{G}_2$) \citep{lindgren2011},
\[
\bm{Q} = \tau^2 \left(\kappa^4\bm{C} + 2\kappa^2\bm{G}_1 + \bm{G}_2 \right),
\]
where $\kappa$ is the Mat\'ern decorrelation rate as before.
We can calculate the marginal variance of the Mat\'ern random field as $\sigma^2 = \left (4 \pi \tau^2 \kappa^2 \right)^{-1}$.

@@

We can extend our model to accommodate spatiotemporal data by adding Gaussian random fields for each time step $t$, $\bm{\epsilon}_{t}$
\[
\begin{aligned}
\mathbb{E}[y_{\bm{s},t}] &= \mu_{\bm{s},t},\\
\mu_{\bm{s},t} &=
g^{-1} \left( \bm{X}_{\bm{s},t} \bm{\beta} + \omega_{\bm{s}} + \epsilon_{\bm{s},t} \right),\\
\bm{\omega} &\sim \operatorname{MVNormal} \left( \bm{0}, \bm{\Sigma}_\omega \right),\\
\bm{\epsilon}_{t} &\sim \operatorname{MVNormal} \left( \bm{0}, \bm{\Sigma}_{\epsilon} \right).
\end{aligned}
\]
In this case we assume the spatiotemporal random fields are independent at each time step, but we could alternatively assume they are structured as a random walk or autoregressive process (demonstrated in Section~\ref{sec:dogar1}).
The spatiotemporal random fields represent latent variables causing spatial correlation that changes with each time step.

@@

In practice, the above models can become considerably more complex by, for example, letting coefficients vary though time, letting coefficients vary through space \citep{hastie1993} as random fields, or adding random intercepts by grouping factors (Figure~\ref{fig:diagram}).
All these components are additive in link space.
Adopting the notation ``main'' for main effects, ``tvc'' for time-varying coefficients, and ``svc'' for spatially varying coefficients, a more complex model could be written as
\[
\begin{aligned}
\mu_{\bm{s},t} &=
g^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \bm{\gamma_t} +
\bm{X}^{\mathrm{svc}}_{\bm{s},t} \zeta_{\bm{s}} +
\alpha_g +
O_{\bm{s},t} +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t} \right).
\end{aligned}
\]
where the $\bm{X}$ represent model matrices, $\bm{\gamma_t}$ represents a vector of coefficients that are allowed to vary through time as random walks or AR(1) processes, $\zeta_{\bm{s}}$ represents spatially varying coefficients following a random field, $\alpha_g$ represents IID random intercepts by group $g$ ($\alpha_g \sim \operatorname{Normal} \left(0, \sigma_\alpha^2 \right)$), $O$ represents an offset variable \citep[][p.~206]{mccullagh1989} (e.g., log sampling effort), and $\omega_{\bm{s}}$ and $\epsilon_{\bm{s},t}$ represent spatial and spatiotemporal intercept random fields as before (Figure~\ref{fig:diagram}).
We demonstrate these model components in the examples below.
% TODO: add MVN(0, sigma)s

@@

% This process component is combined with an observation error family (e.g.,
% Gaussian, gamma, binomial, Tweedie) and link function \(g\) (e.g., identity,
% log, logit) as in any generalized linear model (GLM) \citep{mccullagh1989}. Some
% families can be combined into a two-part ``delta'' or ``hurdle'' model
% \citep{aitchison1955} to model the zero vs.~non-zero observations separately
% from the positive observations. See the model description vignette for an
% expanded description of the statistical model
% (\texttt{browseVignettes("sdmTMB")} or see the
% \href{https://pbs-assess.github.io/sdmTMB/articles/}{rendered version}).

@@

\section{User interface}\label{sec:software}

@@

The \sdmTMB\ package is designed to be both modular and familiar to users of widely used \R\ packages (e.g., \pkg{glmmTMB}, \pkg{lme4}, \pkg{mgcv}).
The user starts by constructing a triangulation mesh for the SPDE approach with \texttt{make\_mesh()} (Figure~\ref{fig:flowchart}).
\texttt{make\_mesh()} is a wrapper for \pkg{fmesher} \citep{fmesher} triangulation mesh functions and users can also construct any mesh with \pkg{fmesher} or \INLA\ and pass it to \texttt{make\_mesh()}.

Fitting is accomplished with \texttt{sdmTMB()}, which has arguments similar to \pkg{glmmTMB}'s \texttt{glmmTMB()} but with additional arguments for how any spatial and spatiotemporal random fields are structured, what column defines time, any time-varying formulas, and any spatially varying formulas.
Observation distributions and links are specified with standard \R\ family functions (e.g., \texttt{binomial()}) plus several \sdmTMB-specific families not available in the \R\ stats library (e.g., \texttt{nbinom2()}, \texttt{delta\_lognormal()}, see \texttt{?sdmTMB::Families}).

The \sdmTMB\ formula syntax (\texttt{formula} argument) follows a standard format that is similar to \pkg{glmmTMB}, \pkg{lme4}, and \pkg{mgcv}.
In addition to standard main effects, the user can include random intercepts (e.g., \texttt{+ (1 | group)}), threshold shaped hockey-stick models \texttt{breakpt()} \citep{barrowman2000}, logistic functions \texttt{logistic()}, and penalized smoothers \texttt{s()} for generalized additive models, GAMS \citep{wood2017a}.
Penalized smoothers use the same \texttt{s()} and \texttt{t2()} syntax as in \pkg{mgcv} \citep{wood2017a}.
Supported functionality includes bivarate smoothers \texttt{s(x, y)},
smoothers varying by continuous or categorical variables \texttt{s(x1, by = x2)},
cyclical smoothers \texttt{s(x, bs = "cc")},
and smoothers with specified basis dimensions \texttt{s(x, k = 4)} \citep{wood2017a}.
Beyond the main formula, \texttt{sdmTMB()} accepts one-sided formulas for coefficients that should vary through time (\texttt{time\_varying}) according to a random walk or AR(1) process (\texttt{time\_varying\_type}) or vary through space as random fields (\texttt{spatial\_varying}).

Once the user makes a call to \texttt{sdmTMB()}, input data structures for a \TMB\ model are constructed internally (Figure~\ref{fig:flowchart}).
If needed, data structures required to implement penalized smoothers are constructed with \texttt{smooth2random()} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
\texttt{sdmTMB()} formats data, establishes parameter starting values, and constructs an objective function with derivatives based on a compiled \proglang{C++} template written for \TMB.
The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016}.
The negative marginal log likelihood is minimized via the non-linear optimization routine \texttt{stats::nlminb()} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \texttt{stats::optimHess()} \citep{gay1990, r2021}.
After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to samples from the joint posterior distribution for Bayesian inference.

A fitted model summary can be viewed with \texttt{print()} or \texttt{summary()} and a set of basic ``sanity'' checks can be run with \texttt{sanity()}.
\texttt{tidy()} returns parameter estimates in standard data frame formats similar to \pkg{broom} \citep{broom}.
Others standard S3 methods are also available such as \texttt{fixef()}, \texttt{confint()}, and \texttt{vcov()}.

Prediction on fitted or new data is accomplished with \texttt{predict()} (\texttt{?predict.sdmTMB}).
The \texttt{predict.sdmTMB()} method is flexible and includes the option to specify a new data frame (\texttt{newdata}),
whether to return predictions on the link or response scale (type),
whether to return standard errors (\texttt{se\_fit},
can be slow if conditioned on random fields),
whether to condition on the random fields (\texttt{re\_form}),
whether to condition on the random intercepts (\texttt{re\_form\_iid}),
which delta model linear predictor to use (\texttt{model}),
whether to take draws from the joint parameter precision matrix (\texttt{nsim}),
and whether to use MCMC samples from a \pkg{tmbstan} model fit (\texttt{mcmc\_samples}).

A variety of model evaluation tools are available.
A \texttt{residuals()} method calculates various types of residuals.
New observations can be simulated from a fitted model with \texttt{simulate()} or new data can be simulated without a fitted model (``de novo'') with \texttt{sdmTMB\_simulate()}.
\texttt{sdmTMB\_cv()} facilitates cross validation.

% Derived quantities can be calculated on the fitted model.
% TODO: more COG refs
% Common quantities of interest include summarizing shifts in the center of gravity (\texttt{get\_cog()}) \citep[e.g.,][]{thorson2016cog}, or tracking total densities through time (\texttt{get\_index()}).

% The \texttt{get\_index()} function generates a weighted average of densities for the prediction grid, integrating across space; for prediction grids with equally sized cells, each cell will contribute equally to the total average, however cells may be of unequal size.

% Data may be simulated with or without a fitted model (\texttt{simulate.sdmTMB()}, \texttt{sdmTMB\_simulate()} respectively).
% Estimation is done via maximum marginal likelihood with the \texttt{sdmTMB()} function, and the resulting fitted model can be used to make predictions (\texttt{predict.sdmTMB()}) or calculate residuals (\texttt{residuals.sdmTMB()}), derive additional quantities of interest, and simulate Bayesian posterior distributions.
% Common quantities of interest may include summarizing shifts in the center of gravity (\texttt{get\_cog()}) \citep[e.g.,][]{thorson2016cog}, or tracking total densities through time (\texttt{get\_index()}).
% The \texttt{get\_index()} function generates a weighted average of densities for the prediction grid, integrating across space; for prediction grids with equally sized cells, each cell will contribute equally to the total average, however cells may be of unequal size.

\sdmTMB\ models can include penalized likelihoods by assigning priors (penalties) to model parameters through the \texttt{sdmTMB()} \texttt{prior} argument (\texttt{?sdmTMBpriors}).
These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
Following other recent SPDE implementations in \TMB\ \citep{breivik2021, osgood-zimmerman2021}, penalized complexity (PC) priors \citep{simpson2017, fuglstad2019} (\texttt{?pc\_matern}) can constrain the spatial range and variance parameters.
These priors or penalties are available both with maximum likelihood estimation and with MCMC sampling.
If a model will be passed to \pkg{tmbstan} with priors, a \texttt{bayesian} logical argument should be set to \texttt{TRUE} to enable Jacobian adjustments for changes of variables (priors applied to parameters that are internally transformed) \citep{carpenter2017}.

@@

% The \sdmTMB\ model is fit by maximum marginal likelihood. Internally,
% a \TMB\ \citep{kristensen2016} model template is used to calculate the
% marginal log likelihood and its gradient, and the negative log likelihood is
% minimized via the non-linear optimization routine \texttt{stats::nlminb()} in
% \R\ \citep{gay1990, r2021}. Random effects are estimated at values that
% maximize the log likelihood conditional on the estimated fixed effects and are
% integrated over via the Laplace approximation \citep{kristensen2016}. After
% rapid model exploration with maximum likelihood, one can optionally pass an
% \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017,
% rstan} via \pkg{tmbstan} \citep{monnahan2018} to estimate the joint posterior
% distribution for Bayesian inference.

@@

We can implement this model with \texttt{sdmTMB()}:

>>>>

Data are often collected in space or in space repeatedly over time. While such data are a rich source of information across many fields,
they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart due to measured and unmeasured variables \citep{cressie1993, diggle2007, cressie2011}.
While measured variables can be accounted for with predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables can cause residual spatial correlation.
Accounting for this residual correlation is important because doing so allows for valid statistical inference \citep{legendre1989a, dormann2007}, can improve predictions \citep[e.g.,][]{shelton2014}, and can be of interest itself \citep[e.g.,][]{thorson2019d, barnett2021}.

Geostatistical generalized linear mixed effects models (GLMMs) with spatially correlated random effects constitute an appropriate class of models for such data \citep{rue2005gmrf, diggle2007, cressie2011}.
Similarly to how random intercepts can account for correlation among groups, spatial or spatiotemporal random effects can account for unmeasured variables that cause observations to be correlated in space or both space and time.
A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed to be drawn from a multivariate normal distribution, constrained by specific covariance functions such as the exponential or Matérn \citep{cressie1993, chiles1999, diggle2007}.
% TODO: decide on using GRF throughout

@@

A basic GLMM with spatial Gaussian random fields can be written as
\[
\begin{aligned}
\mathbb{E}[y_{\bm{s}}] &= \mu_{\bm{s}},\\
\mu_{\bm{s}} &=
g^{-1} \left( \eta_{\bm{s}} \right),\\
\eta_{\bm{s}} &= \bm{X}_{\bm{s}} \bm{\beta} + \omega_{\bm{s}},\\
\end{aligned}
\]
where the expected value $\mathbb{E}[.]$ of an observation $y$ at coordinates in space $\bm{s}$ is equal to mean $\mu_s$.
That mean is the result of an inverse link function $g^{-1}$ applied to a linear predictor $\eta$.
In this case, that linear predictor is the combination of a model matrix $\bm{X}_{\bm{s}}$ multiplied by a vector of coefficients $\bm{\beta}$ and a value from a spatial random field $\omega_{\bm{s}}$.
This spatial random field represents the effect of latent spatial variables that are not otherwise accounted for in the model.
Alternatively, $\omega_{\bm{s}}$ can be thought of representing spatially correlated noise.
More simply, the vector of $\omega_{\bm{s}}$ ($\bm{\omega}$) represents a ``wiggly'' surface with an expected value of zero that is added to the linear predictor in link space (e.g., Figure~\ref{fig:matern-range}).
The vector $\bm{\omega}$ is assumed drawn from a multivariate normal distribution with a covariance matrix $\boldsymbol{\Sigma}_\omega$,
\[
\boldsymbol{\omega} \sim \operatorname{MVNormal}
  \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\omega \right),
\]
constrained by some function that defines the rate at which spatial covariance decays with distance.

@@

Various covariance functions are possible, but a popular and flexible one is the Mat\'ern \citep{whittle1954, matern1986} (Figure~\ref{fig:matern-range}).
If we define $\lVert h \rVert$ as the Euclidean distance between spatial points $\bm{s}_j$ and $\bm{s}_k$, we can represent the Mat\'ern covariance $\Phi$ as
\[
\Phi\left( \bm{s}_j, \bm{s}_k \right) =
  \frac{\sigma^2}{2^{\nu - 1}\Gamma(\nu)}
  \left(\kappa \lVert h \rVert \right)^\nu K_\nu
  \left( \kappa \lVert h \rVert \right).
\]
The parameter $\sigma^2$ is the marginal variance (magnitude of the random field ``wiggles''), $\Gamma$ represents the Gamma function, $K_\nu$ represents the modified Bessel function of the second kind, and $\kappa$ represents the spatial decorrelation rate.
The parameter $\nu$ controls the smoothness of the covariance function.
In practice, $\nu$ is challenging to estimate and herein is fixed at $\nu = 1$ \citep{lindgren2011}.
A more interpretable parameter than the spatial decorrelation rate $\kappa$ is the spatial range---a distance at which two points are effectively independent.
A common definition is $\sqrt{8 \nu} / \kappa$ (so if $\nu = 1$, $\mathrm{range} = \sqrt{8} / \kappa$), which has the empirically derived property of the distance at which correlation decays to $\approx 0.13$ \citep{lindgren2011} (Figure~\ref{fig:matern-range}).

@@

In practice, working with the dense covariance matrix $\bm{\Sigma}$ is computationally expensive
and methods for working directly with its inverse, the precision matrix $\bm{Q}$, are more efficient ($\bm{Q} = \bm{\Sigma}^{-1}$) \citep{simpson2012}.
One such approach is the SPDE approach.
A full description of the SPDE approach is beyond the scope of this paper.
Instead, we refer to the following literature:
\citet{lindgren2011} develops the approach.
\citet{lindgren2015} and \cite{bakka2018review} summarize the SPDE approach for spatial modelling in the context of \INLA.
The second chapter of \citet{krainski2018} provides an overview of the SPDE approach to spatial modelling with a focus on linking theory to code.
\citet{miller2019} contains an introduction to the SPDE and illustrates its equivalence to penalized smoothing approaches.
\citet{lindgren2022ten} provides a recent review of the approach and its applications over the last decade.

For a user of \sdmTMB, the following are the important elements to understand.
First, the SPDE approach links Gaussian random fields (GRFs) that have a Mat\'ern covariance function to Gaussian \emph{Markov} random fields (GMRFs) in such a way that a GMRF can be a good approximation to a GRF \citep{lindgren2011}.
This means that GRF models can be computationally fit as GMRFs.
By working with GMRFs, one can take advantage of theory developed to estimate their sparse precision matrix efficiently \citep{rue2005gmrf, lindgren2011} and avoid the inversion of large dense matrices.
Second, the SPDE involves piece-wise linear basis functions that are defined by a triangulation over the spatial area of interest \citep{lindgren2011}.
Commonly, this is referred to as a ``mesh''.
The properties of this mesh (e.g., its resolution and how far it extends beyond the data) affect the accuracy and computational efficiency of the SPDE approach \citep{lindgren2011}.
Third, working with the SPDE approach and the precision matrix of a GMRF ($\bm{Q}$) introduces an alternative parameter $\tau$, which scales the precision matrix, and involves three sparse matrices associated with the mesh ($\bm{C}$, $\bm{G}_1$, and $\bm{G}_2$) \citep{lindgren2011},
\[
\bm{Q} = \tau^2 \left(\kappa^4\bm{C} + 2\kappa^2\bm{G}_1 + \bm{G}_2 \right),
\]
where $\kappa$ is the Mat\'ern decorrelation rate as before.
We can calculate the marginal variance of the Mat\'ern random field as $\sigma^2 = \left (4 \pi \tau^2 \kappa^2 \right)^{-1}$.

@@

We can extend our spatial model to accommodate spatiotemporal data by adding Gaussian random fields for each time step $t$, $\bm{\epsilon}_{t}$
\[
\begin{aligned}
\mathbb{E}[y_{\bm{s},t}] &= \mu_{\bm{s},t},\\
\mu_{\bm{s},t} &=
g^{-1} \left( \bm{X}_{\bm{s},t} \bm{\beta} + \omega_{\bm{s}} + \epsilon_{\bm{s},t} \right),\\
\bm{\omega} &\sim \operatorname{MVNormal} \left( \bm{0}, \bm{\Sigma}_\omega \right),\\
\bm{\epsilon}_{t} &\sim \operatorname{MVNormal} \left( \bm{0}, \bm{\Sigma}_{\epsilon} \right).
\end{aligned}
\]
In this case, we assume the spatiotemporal random fields are independent at each time step, but we could alternatively assume they are structured as a random walk or autoregressive process (demonstrated in Section~\ref{sec:dogar1}).
The spatiotemporal random fields represent latent variables causing spatial correlation that changes with each time step.

@@

In practice, the above models can become considerably more complex by, for example, letting coefficients vary though time, letting coefficients vary through space \citep{hastie1993} as random fields, or adding random intercepts by grouping factors (Figure~\ref{fig:diagram}).
All these components are additive in link space.
Adopting the notation ``main'' for main effects, ``tvc'' for time-varying coefficients, and ``svc'' for spatially varying coefficients, a more complex model could be written as
\[
\begin{aligned}
\mu_{\bm{s},t} &=
g^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \bm{\gamma_t} +
\bm{X}^{\mathrm{svc}}_{\bm{s},t} \zeta_{\bm{s}} +
\alpha_g +
O_{\bm{s},t} +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t} \right).
\end{aligned}
\]
where the $\bm{X}$ represent model matrices, $\bm{\gamma_t}$ represents a vector of coefficients that are allowed to vary through time as random walks or AR(1) processes, $\zeta_{\bm{s}}$ represents spatially varying coefficients following a random field, $\alpha_g$ represents IID random intercepts by group $g$ ($\alpha_g \sim \operatorname{Normal} \left(0, \sigma_\alpha^2 \right)$), $O_{\bm{s},t}$ represents an offset variable \citep[][p.~206]{mccullagh1989} (e.g., log sampling effort), and $\omega_{\bm{s}}$ and $\epsilon_{\bm{s},t}$ represent spatial and spatiotemporal intercept random fields as before (Figure~\ref{fig:diagram}).
We demonstrate these model components in Sections~\ref{pcod}, \ref{dogfish}, and \ref{svc-owls}.
% TODO: add MVN(0, sigma)s

@@

% This process component is combined with an observation error family (e.g.,
% Gaussian, gamma, binomial, Tweedie) and link function \(g\) (e.g., identity,
% log, logit) as in any generalized linear model (GLM) \citep{mccullagh1989}. Some
% families can be combined into a two-part ``delta'' or ``hurdle'' model
% \citep{aitchison1955} to model the zero vs.~non-zero observations separately
% from the positive observations. See the model description vignette for an
% expanded description of the statistical model
% (\code{browseVignettes("sdmTMB")} or see the
% \href{https://pbs-assess.github.io/sdmTMB/articles/}{rendered version}).

@@

\section{Software design and user interface}\label{sec:software}

@@

The \sdmTMB\ package is designed to be both modular and familiar to users of widely used \R\ packages (e.g., \pkg{glmmTMB}, \pkg{lme4}, \pkg{mgcv}).
The user starts by constructing a triangulation mesh for the SPDE approach with \fct{make\_mesh()} (Figure~\ref{fig:flowchart}).
\fct{make\_mesh()} is a wrapper for \pkg{fmesher} \citep{fmesher} triangulation mesh functions and users can also construct any mesh with \pkg{fmesher} or \INLA\ and pass it to \fct{make\_mesh()}.

Fitting is accomplished with \fct{sdmTMB()}, which has arguments similar to \pkg{glmmTMB}'s \fct{glmmTMB()} but with additional arguments for how any spatial and spatiotemporal random fields are structured, what column defines time, any time-varying formulas, and any spatially varying formulas.
Observation distributions and links are specified with standard \R\ family functions (e.g., \fct{binomial()}) plus several \sdmTMB-specific families not available in the \R\ stats library (e.g., \fct{nbinom2()}, \fct{delta\_lognormal()}, see \fct{?sdmTMB::Families}).

The \sdmTMB\ formula syntax (\code{formula} argument) follows a standard format that is similar to \pkg{glmmTMB}, \pkg{lme4}, and \pkg{mgcv}.
In addition to standard main effects, the user can include random intercepts (e.g., \code{+ (1 | group)}), threshold shaped hockey-stick models \fct{breakpt()} \citep{barrowman2000}, logistic functions \fct{logistic()}, and penalized smoothers \fct{s()} for generalized additive models, GAMS \citep{wood2017a}.
Penalized smoothers use the same \fct{s()} and \fct{t2()} syntax as in \pkg{mgcv} \citep{wood2017a}.
Supported functionality includes bivarate smoothers \fct{s(x, y)},
smoothers varying by continuous or categorical variables \fct{s(x1, by = x2)},
cyclical smoothers \fct{s(x, bs = "cc")},
and smoothers with specified basis dimensions \fct{s(x, k = 4)} \citep{wood2017a}.
Beyond the main formula, \fct{sdmTMB()} accepts one-sided formulas for coefficients that should vary through time (\code{time\_varying}) according to a random walk or AR(1) process (\code{time\_varying\_type}) or vary through space as random fields (\code{spatial\_varying}).

Once the user makes a call to \fct{sdmTMB()}, input data structures for a \TMB\ model are constructed internally (Figure~\ref{fig:flowchart}).
If needed, data structures required to implement penalized smoothers are constructed with \fct{smooth2random()} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
\fct{sdmTMB()} formats data, establishes parameter starting values, and constructs an objective function with derivatives based on a compiled \proglang{C++} template written for \TMB.
The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016}.
The negative marginal log likelihood is minimized via the non-linear optimization routine \fct{stats::nlminb()} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \fct{stats::optimHess()} \citep{gay1990, r2021}.
After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to samples from the joint posterior distribution for Bayesian inference.

A fitted model summary can be viewed with \fct{print()} or \fct{summary()} and a set of basic ``sanity'' checks can be run with \fct{sanity()}.
\fct{tidy()} returns parameter estimates in standard data frame formats similar to \pkg{broom} \citep{broom}.
Others standard S3 methods are also available such as \fct{fixef()}, \fct{confint()}, and \fct{vcov()}.

Prediction on fitted or new data is accomplished with \fct{predict()} (\fct{?predict.sdmTMB}).
The \fct{predict.sdmTMB()} method is flexible and includes the option to specify a new data frame (\fct{newdata}),
whether to return predictions on the link or response scale (type),
whether to return standard errors (\code{se\_fit},
can be slow if conditioned on random fields),
whether to condition on the random fields (\code{re\_form}),
whether to condition on the random intercepts (\code{re\_form\_iid}),
which delta model linear predictor to use (\code{model}),
whether to take draws from the joint parameter precision matrix (\code{nsim}),
and whether to use MCMC samples from a \pkg{tmbstan} model fit (\code{mcmc\_samples}).

A variety of model evaluation tools are available.
A \fct{residuals()} method calculates various types of residuals.
New observations can be simulated from a fitted model with \fct{simulate()} or new data can be simulated without a fitted model (``de novo'') with \fct{sdmTMB\_simulate()}.
\fct{sdmTMB\_cv()} facilitates cross validation.

% Derived quantities can be calculated on the fitted model.
% TODO: more COG refs
% Common quantities of interest include summarizing shifts in the center of gravity (\fct{get\_cog()}) \citep[e.g.,][]{thorson2016cog}, or tracking total densities through time (\fct{get\_index()}).

% The \fct{get\_index()} function generates a weighted average of densities for the prediction grid, integrating across space; for prediction grids with equally sized cells, each cell will contribute equally to the total average, however cells may be of unequal size.

% Data may be simulated with or without a fitted model (\fct{simulate.sdmTMB()}, \fct{sdmTMB\_simulate()} respectively).
% Estimation is done via maximum marginal likelihood with the \fct{sdmTMB()} function, and the resulting fitted model can be used to make predictions (\fct{predict.sdmTMB()}) or calculate residuals (\fct{residuals.sdmTMB()}), derive additional quantities of interest, and simulate Bayesian posterior distributions.
% Common quantities of interest may include summarizing shifts in the center of gravity (\fct{get\_cog()}) \citep[e.g.,][]{thorson2016cog}, or tracking total densities through time (\fct{get\_index()}).
% The \fct{get\_index()} function generates a weighted average of densities for the prediction grid, integrating across space; for prediction grids with equally sized cells, each cell will contribute equally to the total average, however cells may be of unequal size.

\sdmTMB\ models can include penalized likelihoods by assigning priors (penalties) to model parameters through the \fct{sdmTMB()} \code{prior} argument (\code{?sdmTMBpriors}).
These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
Following other recent SPDE implementations in \TMB\ \citep{breivik2021, osgood-zimmerman2021}, penalized complexity (PC) priors \citep{simpson2017, fuglstad2019} (\code{?pc\_matern}) can constrain the spatial range and variance parameters.
These priors or penalties are available both with maximum likelihood estimation and with MCMC sampling.
If a model will be passed to \pkg{tmbstan} with priors, a \code{bayesian} logical argument should be set to \code{TRUE} to enable Jacobian adjustments for changes of variables (priors applied to parameters that are internally transformed) \citep{carpenter2017}.

@@

% The \sdmTMB\ model is fit by maximum marginal likelihood. Internally,
% a \TMB\ \citep{kristensen2016} model template is used to calculate the
% marginal log likelihood and its gradient, and the negative log likelihood is
% minimized via the non-linear optimization routine \code{stats::nlminb()} in
% \R\ \citep{gay1990, r2021}. Random effects are estimated at values that
% maximize the log likelihood conditional on the estimated fixed effects and are
% integrated over via the Laplace approximation \citep{kristensen2016}. After
% rapid model exploration with maximum likelihood, one can optionally pass an
% \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017,
% rstan} via \pkg{tmbstan} \citep{monnahan2018} to estimate the joint posterior
% distribution for Bayesian inference.

@@

We can implement this model with \fct{sdmTMB()}:
